---
title: "Summarizing sets of radiocarbon dates using baydem's standard pipeline"
author: "Michael Holton Price"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarizing sets of radiocarbon dates using baydem's standard pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview
`baydem` is an R package for doing end-to-end inference on radiocarbon dates in order to summarize their distribution across time. The majority of analyses involve the same exact steps, so `baydem` provides an easy-to-use standard pipeline for doing a typical analysis. The standard pipeline differs slightly when running a simulation as opposed to running using existing data saved in a .csv file. The next section describes the standard pipeline for a simulation and the following section describes the standard pipeline using existing data. There are two key variables that are used by all the standard pipeline functions: `data_dir` and `analysis_name`. `data_dir` is a directory in which to store analysis data. `analysis_name` is a unique name for a given analysis in `data_dir`. Multiple analyses can be stored in `data_dir`. For this vignette, we will use the working directory to store data:

```{r}
data_dir <- getwd()
```

# The standard pipeline for a simulation
This section of the vignette describes the standard pipeline for a simulation, which consists of:

## Step 0. Create simulated data
In order to demonstrate the steps, we need a starting .csv file containing radiocarbon measurements. Create such a file from simulated data. In practice, the user will create the file from real radiocarbon measurements.

```{r}
# Load baydem
library(baydem)
# Simulate data with a truncated Gaussian mixture for the density model. For
# details on this step (for example, on the error_spec), see the documentation
# for simulate_rc_data. Make N=100 simulated samples.

# The simulation parameter vector (un-truncated) with K=2 mixtures components.
# pi1/pi2 are the mixture proportions, mu1/mu2 the mixture means (in years AD,
# not BP), and s1/s2 the mixture standard deviations (in years).
th_sim <-
  c(
    pi1 = 0.2,
    pi2 = 0.8,
    mu1 = 775,
    mu2 = 1000,
    s1 = 35,
    s2 = 45
  )

sim <- simulate_rc_data(
                        list(
                             model_spec = list(density_type="trunc_gauss_mix",
                                               th=c(th_sim,600,1300),
                                               error_spec=list(
                                                               type="unif_fm",
                                                               min=.0021,
                                                               max=.0028
                                                              ),
                                               is_AD=T
                                              ),
                             N=100,
                             calib_curve="intcal20",
                             seed=1002
                            )
                       )

```


## Step 1. Set the simulated data for the analysis
Before setting the simulated data (and if necessary), delete the analysis file that `set_sim` will create (otherwise, an error will be thrown).

```{r}
sim_save_file <- file.path(data_dir, "sim_example.rds")
if (file.exists(sim_save_file)) {
  success <- file.remove(sim_save_file)
}
```

Set the analysis name and call `set_sim`.

```{r}
analysis_name <- "sim_example"
set_sim(data_dir,analysis_name,sim)
```

`set_sim` saves an R list object to the file sim_example.rds with three named elements:
```{r}
analysis <- readRDS(sim_save_file)
print(names(analysis))

```

`rc_meas` contains the radiocarbon measurements (see the documentation for `import_rc_data`), `calib_df` contains the calibration curve information (see the documentation for `load_calib_curve`), and `sim` is the sim object that was input.

## Step 2: Determine the range of calendar dates to allow during Bayesian inference
Call `calc_tau_range` to determine the range of calendar dates that span the radiocarbon measurements.

```{r}
tau_range <- calc_tau_range(analysis$rc_meas,
                            calibration_curve="intcal20",
                            dtau=5)
```

`tau_range` is a list with `tau_min` and `tau_max` (the minimum maximum calendar dates). By setting `dtau` to 5, the ranges are specified to multiples of 5 years (using floor for the minimum and ceiling for the maximum), which is the spacing of the IntCal20 calibration curve.


```{r}
print(tau_range)
```

## Step 3: Specify the parametric model to use for Bayesian inference
Use `set_density_model` to set the parametric model to be used for Bayesian inference. Currently, baydem only supports a truncated Gaussian mixture, though other parameteric models will likely be added in the future. The truncation limits were calculated in the preceding step. Although we know that the simulated data were made with $K=2$ mixtures components, with non-simulated data one does not know ahead of time how many mixture components to use. By setting $K$ to a vector with 2, 3, and 4 mixture components at the next step, Bayesian inference will be done for each number of mixture components below.

```{r}
density_model <- list(type="trunc_gauss_mix",
                      tau_min=tau_range$tau_min,
                      tau_max=tau_range$tau_max,
                      K=2:4)
set_density_model(data_dir,analysis_name,density_model)
```

"sim_example.rds" now contains the density model:

```{r}
analysis <- readRDS(sim_save_file)
print(names(analysis))
```

## Step 4: Do Bayesian inference
Call `do_bayesian_inference` to do Bayesian inference (using `sample_theta`) for each of the number of mixture components specified in the preceding step (2, 3, and 4). The hyperparameters (hp) specify the prior probabilities for the parameter vector, as well as the spacing to use for the sampling grid, which is used in the Riemann sum that approximates the integral required for the likelihood calculation. Unlike with using an existing dataset, for simulations it is uncessary to use `set_calib_curve` since the calibration curve was already set in the Step 1 above. `do_bayesian_inference` will probably take between 10 and 20 minutes to finish.

```{r}
hp <-
  list(
    # Parameter for the dirichlet draw of the mixture probabilities
    alpha_d = 1,
    # The gamma distribution shape parameter for sigma
    alpha_s = 10,
    # The gamma distribution rate parameter for sigma, yielding a mode of 50
    alpha_r = (10 - 1) / 50,
    # Spacing for the measurement matrix (years)
    dtau = 1
  )
do_bayesian_inference(data_dir,analysis_name,hp)
```

"sim_example.rds" now contains the Bayesian inference information:

```{r}
analysis <- readRDS(sim_save_file)
print(names(analysis))
```

## Step 5: Do Bayesian summary
Call `do_bayesian_summary` to calculate summary information for the best model from the preceding step. The best model is the one with the lowest WAIC value among the different models with different numbers of mixture conponents (2, 3, and 4).

```{r}
do_bayesian_summary(data_dir,analysis_name)
```

"sim_example.rds" now contains the Bayesian summary information:

```{r}
analysis <- readRDS(sim_save_file)
print(names(analysis))
```

## Step 6: Plot best model
Call `plot_best_solution` to plot the best solution (best WAIC). Since this is a simulation, a known density can be added, so use `add_known_density=TRUE`.


```{r, fig.width=8, fig.height=6}
plot_best_solution(data_dir,analysis_name,add_known_density=TRUE)
```

# The standard pipeline for a simulation
This section of the vignette describes the standard pipeline for an analysis that starts with radiocarbon measurements stored in a .csv file. First, write the simulated data already created to a .csv file. This creates the type of file that will typically be input for an analysis, albeit from simulated data. In practice, the user will create the file from real radiocarbon measurements.

```{r}
output_df <- data.frame(rc_year_bp=sim$data$rc_meas$trc_m,
                        rc_year_bp_error=sim$data$rc_meas$sig_trc_m)
print(head(output_df))
write.csv(output_df, "rc_data.csv", row.names=FALSE)
```

## Step 0. Load the radiocarbon data
Load the radiocarbon data using `import_rc_data`. Since the first column of the .csv file contains the radiocarbon year values (years before present in radiocarbon years) and the second column contains the corresponding uncertanties, column names do not need to be specified. If the pertinent data are in different columns, see the documentation for `import_rc_data` for alternative ways to load the radicarbon data.


```{r}
rc_meas <- import_rc_data("rc_data.csv")
print(names(rc_meas))
print(head(rc_meas$trc_m))
print(head(rc_meas$trc_m))
```

## Step 1. Set the radiocarbon data for the analysis
Before setting the radiocarbon data data (and if necessary), delete the analysis file that `set_rc_data` will create (otherwise, an error will be thrown).

```{r}
rc_save_file <- file.path(data_dir, "rc_example.rds")
if (file.exists(rc_save_file)) {
  success <- file.remove(rc_save_file)
}
```

Next, call `set_rc_meas`.

```{r}
analysis_name <- "rc_example"
set_rc_meas(data_dir,analysis_name,rc_meas)
```

After setting the radiocarbon measurments, the analysis file contains a list with one named element, `rc_meas` (the radiocarbon measurements):
```{r}
analysis <- readRDS(rc_save_file)
print(names(analysis))
```

## Step 2: Determine the range of calendar dates to allow during Bayesian inference
Call `calc_tau_range` to determine the range of calendar dates that span the radiocarbon measurements.

```{r}
tau_range <- calc_tau_range(analysis$rc_meas,
                            calibration_curve="intcal20",
                            dtau=5)
```

## Step 3: Specify the parametric model to use for Bayesian inference
Use `set_density_model` to set the parametric model to be used for Bayesian inference. This step is identical to that for a simulation (see above).

```{r}
density_model <- list(type="trunc_gauss_mix",
                      tau_min=tau_range$tau_min,
                      tau_max=tau_range$tau_max,
                      K=2:4)
set_density_model(data_dir,analysis_name,density_model)
```

"rc_example.rds" now contains the density model:

```{r}
analysis <- readRDS(rc_save_file)
print(names(analysis))
```

## Step 4: Set the calibration curve to use for subsequent steps
Use `set_calib_curve` to set the calibration curve. The third input can be either the name of a calibration curve or a data frame that specifies the calibration curve (see `load_calib_curve` for details on the options and formats).

```{r}
set_calib_curve(data_dir,analysis_name,"intcal20")
```

"rc_example.rds" now contains the calibration data frame:

```{r}
analysis <- readRDS(rc_save_file)
print(names(analysis))
```

## Step 5: Do Bayesian inference
Call `do_bayesian_inference` to do Bayesian inference (using `sample_theta`). This step is identical to that for a simulation (see above).

```{r}
hp <-
  list(
    # Parameter for the dirichlet draw of the mixture probabilities
    alpha_d = 1,
    # The gamma distribution shape parameter for sigma
    alpha_s = 10,
    # The gamma distribution rate parameter for sigma, yielding a mode of 50
    alpha_r = (10 - 1) / 50,
    # Spacing for the measurement matrix (years)
    dtau = 1
  )
do_bayesian_inference(data_dir,analysis_name,hp)
```

"rc_example.rds" now contains the Bayesian inference information:

```{r}
analysis <- readRDS(rc_save_file)
print(names(analysis))
```

## Step 6: Do Bayesian summary
Call `do_bayesian_summary` to calculate summary information for the best model from the preceding step. This step is identical to that for a simulation (see above).

```{r}
do_bayesian_summary(data_dir,analysis_name)
```

"rc_example.rds" now contains the Bayesian summary information:

```{r}
analysis <- readRDS(rc_save_file)
print(names(analysis))
```

## Step 7: Plot best model
Call `plot_best_solution` to plot the best solution (best WAIC).

```{r, fig.width=8, fig.height=6}
plot_best_solution(data_dir,analysis_name)
```

